pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-18 article.tex
Updates from Overleaf

@@

@@

\begin{lstlisting}[
    language=R, floatplacement=ht, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in figure~\ref{fig:yule_tree}
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always & no & JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the first research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tbl:RQ1}
\end{table}

@@

In the previous example, we used a known generative tree prior to match with the inference model.
However, a novel tree prior, by definition, is non-standard, thus cannot be part of a standard inference model (i.e. an inference model using a standard tree prior). 
One option is to choose to hand-pick a related 
standard tree prior, as done in ERCs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.

A second research question that \verb;pirouette; can answer, is:

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    candidate & best candidate & yes & JC, strict, Yule \\
    candidate & best candidate & yes & JC, strict, BD   \\
    ...       & ...            & ... & ...              \\
    candidate & best candidate & yes & GTR, RLN, CCP    \\
    candidate & best candidate & yes & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the second research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
\end{table}

In the second step we set up our experiments. 
Within this ERQ, each of our experiments contain a different
inference model $I$, but has condition $C$ that will only let $I$
be run in inference if that inference model $I$ is the 
best candidate inference model, that is, 
it is the inference model with the highest evidence.
Setting up experiments like this is thought to be so common,
that \verb;pirouette; has a simple function to do:
\verb;create_all_experiments; will create 40 experiments,
containing the inference models of all combinations of 4 site 
models, 2 clock models and 5 tree priors.

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:create_all_experiments},
  caption = Create all 40 candidate experiments
]
experiments <- create_all_experiments()
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lts:create_pir_params_defaults},
  caption = Create a \texttt{pir\_params} with many defaults
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments
)
\end{lstlisting}

@@

A third research question that \verb;pirouette; can answer is:

@@

This time, however, we are interested in creating a twin tree,
as such a twin tree allows us to measure a baseline error produced
by the BEAST2 run.
A twin tree
has the same topology as the given tree, 
yet its branching times are simulated according 
to a tree prior already implemented by BEAST2 to produce a posterior. 
Because for a twin tree we know the tree prior 
and because we use that tree prior in inference, 
we expect BEAST2 to produce a posterior of phylogenies
that is close to the twin tree.
As the inference model for the twin tree is (more) completely
specified than the true tree, 
we expect that a twin tree and its posterior trees are more similar
than a true tree and its posterior trees.
Due to this, the twinning is used a control for the experiment.

@@

A fourth research question that \verb;pirouette; can answer, is:

@@

A fifth research question that \verb;pirouette; can answer, is:

>>>>

\subsection{Question}

@@

\subsection{Answer}

@@

\begin{lstlisting}[
    language=R, floatplacement=ht, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in figure~\ref{fig:yule_tree}.
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always & no & JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 1.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tab:RQ1}
\end{table}

@@

\subsection{Question}

\iffalse
In the previous example, we used a known generative tree prior to match with the inference model.
However, a novel tree prior, by definition, is non-standard, thus cannot be part of a standard inference model (i.e. an inference model using a standard tree prior).
One option is to choose to hand-pick a related 
standard tree prior, as done in ERQs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.
\fi

In the previous example we selected the inference model $\mathit{I}$ to match with a known generative tree prior $\mathit{P_{0}}$.
However, a novel tree prior $\mathit{P}$, by definition, is non-standard, thus cannot be part of a standard inference model (i.e. an inference model using a standard tree prior).
In this case the user has to specify the the tree prior $\mathit{t_{p}}$ to be used in the inference from a list of standard tree priors. In general, having no information on $\mathit{P}$, this choice is arbitrary.
\verb;pirouette; provides two ways to deal with this problem. The choice can be made specifying the inference conditions $\mathit{C}$ to be used.

If the user has sufficient reasons to believe that one standard tree prior could be more suitable than others for the inference, they can set it as part of the inference model and to adopt site and clock models used to simulate the alignment. This option can be specified selecting "model\_type" to be "generative" as part of the inference conditions (as seen in ERQ 1, table~\ref{tab:RQ1}).

If, instead, there is no preference for a specific tree prior, the user can let \verb;pirouette; select the best inference model from a list of candidate models. The model selection is performed automatically based on the criterion of the highest evidence (aka marginal likelihood). This option can be specified providing any number of experiments, selecting "model\_type" to be "candidate" for each of them, and "run\_if" to be "best\_candidate" (table~\ref{tab:RQ2} will summarize this for ERQ 2).

\iffalse
One option is to choose to hand-pick a related 
standard tree prior, as done in ERQs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.
\fi
\iffalse
A second research question that \verb;pirouette; can answer, is:
\fi

We show the procedure using \verb;pirouette; to answer the second research question:

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    candidate & best candidate & yes & JC, strict, Yule \\
    candidate & best candidate & yes & JC, strict, BD   \\
    ...       & ...            & ... & ...              \\
    candidate & best candidate & yes & GTR, RLN, CCP    \\
    candidate & best candidate & yes & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 2.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

\iffalse
In the second step we set up our experiments. 
Within this ERQ, each of our experiments contain a different
inference model $I$, but has condition $C$ that will only let $I$
be run in inference only if that inference model $I$ is the 
best candidate inference model, that is, 
it is the inference model with the highest evidence.
Setting up experiments like this is thought to be so common,
that \verb;pirouette; has a simple function to do:
\verb;create_all_experiments; will create 40 experiments,
containing the inference models of all combinations of 4 site 
models, 2 clock models and 5 tree priors.
\fi

In the second step we set up our experiments. Unlike ERQ 1, here we provide a set of experiments $\{\mathit{X_{i}}\}_{i=1}^{40}$. All the experiments have the same inference conditions $\mathit{C_{i}} = \mathit{C}$ but different candidate inference models $\mathit{I_{i}}$. Yet, only one of them will be used to run the actual inference. According to the specified $\mathit{C}$, \verb;pirouette; will measure the evidence for each of them and select  only the one with the highest evidence as the best candidate.

Setting up experiments like this is thought to be so common,
that \verb;pirouette; has a simple function to do:
\verb;create_all_experiments; will create the entire set of 40 experiments,
containing the inference models of all combinations of 4 site 
models, 2 clock models and 5 tree priors.

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:create_all_experiments},
  caption = Create all 40 candidate experiments.
]
experiments <- create_all_experiments()
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lts:create_pir_params_defaults},
  caption = Create a \texttt{pir\_params} with many defaults.
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments
)
\end{lstlisting}

@@

\subsection{Question}

In ERQ 1 we built the twin pipeline. At the end of the section we discussed how it has not much relevance if considered as a stand-alone process. Here, instead, we will use it as a control for the main pipeline, where the generative tree prior $\mathit{P}$ is unknown, but the site and clock models used for the inference are the ones used to simulate the alignment.
This leads to the third ERQ:

@@

\iffalse
This time, however, we are interested in creating a twin tree,
as such a twin tree allows us to measure a baseline error produced
by the BEAST2 run.
A twin tree
has the same topology as the given tree, 
yet its branching times are simulated according 
to a tree prior already implemented by BEAST2 to produce a posterior. 
Because for a twin tree we know the tree prior 
and because we use that tree prior in inference, 
we expect BEAST2 to produce a posterior of phylogenies
that is close to the twin tree.
As the inference model for the twin tree is (more) completely
specified than the true tree, 
we expect that a twin tree and its posterior trees are more similar
than a true tree and its posterior trees.
Due to this, the twinning is used a control for the experiment.
\fi

This time, however, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branching times are simulated according to the standard tree prior used by BEAST2 to produce the posterior for the main pipeline. 
Since there is coherence between the inference model and the generative model, we expect BEAST2 to produce a posterior of phylogenies more similar to the original one, respect to the outcome of the main pipeline. 
Due to this, the twinning process is used to provide a control for the main experiment.

@@

\subsection{Question}

To answer ERQ 2 we used a evidence-based model selection procedure to obtain an error distribution.
However, it is unclear to what extent this error is due to the mismatch between the tree priors (from generative and inference models) or due to stochasticity. As mentioned in ERQ 1, \verb;pirouette; provides a tool to evaluate this, using the twinning process. We now apply it to the same pipeline followed in ERQ 2.
This leads to our fourth example research question:

@@

\subsection{Question}

In the previous research question we have seen how to use \verb;pirouette; to  run the analysis on main pipeline, either using a 'generative' or a 'best\_candidate' inference model. We also saw that to exclude the baseline error due to stochasticity it is possible to run a parallel twin pipeline. In this way two error distributions are obtained. Comparing them it is possible to estimate the error made by BEAST2 when inferring a posterior distribution starting from a phylogeny generated under an unknown generative tree prior.
We can sum up all \verb;pirouette;'s features presented so far in the fifth research question:
